Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2015-04-02 paint-contour-tree.tex

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
This is clearly an improvement when the tree is short and fat, but it can be even better. 
A calculation yields that for any decomposition of a rooted balanced binary tree, 
our bound is $O(t\alpha(t) + n)$. 
%
%
%In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
%However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time of the algorithm of \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$ time.

>>>>

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
This is clearly an improvement when the tree is short and fat, but it can be even better. 
Consider a contour tree that is constant balanced, meaning that the ratio of sizes of sibling subtrees
is $\Theta(1)$. 
A calculation yields that for any decomposition of such a contour tree, 
our bound is $O(t\alpha(t) + n)$. 
%
%In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
%However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below,
so $t = \Omega(n)$. The running time of the algorithm of \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$ time.
